Lattices of acyclic pipe dreams

### 2023-10-06 acyclicFacets/draft/latticePipeDreams.tex
corrections comments Nantel

To prove this result, we will need the following observation.
Note that throughout the proof, the pipes, rows, and column are all indexed by indices ranging from $0$ to~$n$.

@@

\begin{proof}
Let~$r$ be the row and $c$ be the column of the crossing~$x$ (careful that all indices range from~$0$ to~$n$).
Since $i$ has an elbow southwest of $x$, we have $r \le i-1$.
Since $j$ has an elbow northeast of~$x$, there are at most $r-1 \le i-2$ pipes that cross north of $(c,r)$ in column~$c$.
Hence, at least two pipes smaller than~$i$ reach the northern border before column~$c$.
Excluding the pipe~$0$, there is at least one pipe $h > 0$ such that $h < i < j$ and $\omega^{-1}(h) < \omega^{-1}(j) < \omega^{-1}(i)$.
This is a $132$ pattern in the permutation~$\omega$, therefore $\omega$ is not dominant.
\end{proof}

\begin{proof}[Proof of \cref{prob:nuAcyclicProperty}]
For the forward implication, assume that $\omega$ is not dominant, and consider~${i,j,k \in [n]}$ such that~$i < j < k$ while~$\omega^{-1}(i) < \omega^{-1}(k) < \omega^{-1}(j)$, and such that~$k$ is maximal for this property.
Consider any pipe dream~$P \in \pipeDreams(0\omega)$ obtained by capping any pipe dream of~$\pipeDreams(\omega)$ by a top row of contacts.
Because of this top row of contacts, $\omega^{-1}(u) < \omega^{-1}(v)$ implies~$u \contactLess{P} v$.
We now consider the eastmost point~$x$ of pipe~$k$ along row~$i$, and let~$\ell$ denote the other pipe at this point.
We distinguish two cases:
\begin{enumerate}[(i)]
\item Assume first that~$k$ and~$\ell$ cross at~$x$. Since $\omega^{-1}(k) < \omega^{-1}(\ell)$, we have~$k \contactLess{P} \ell$. Moreover, since~$\ell\ne i$, the pipe~$\ell$ must have an elbow west of~$x$ along row~$i$, which implies that~$\ell \contactLess{P} k$. Hence, we have both~$k \contactLess{P} \ell$ and~$\ell \contactLess{P} k$, so that~$P$ is cyclic.
\item Assume now that~$k$ and~$\ell$ touch at~$x$, so that~$k \contactLess{P} \ell$. If~$\ell < k$, then~$k$ and~$\ell$ must cross before~$x$, and flipping~$x$ in~$P$ yields a cyclic facet. If~$\omega^{-1}(\ell) < \omega^{-1}(k)$, then~$\ell \contactLess{P} k$, so that~$P$ is again cyclic. Hence, we can assume that~$k < \ell$ and~$\omega^{-1}(k) < \omega^{-1}(\ell)$. By maximality of~$k$, we obtain that~$j$ and~$\ell$ do not cross, and thus that~$j$ passes northwest of~$x$. We conclude that~$j \contactLess{P} k$. Since~$\omega^{-1}(k) < \omega^{-1}(j)$, we also have~$k \contactLess{P} j$, so that~$P$ is cyclic.
\end{enumerate}

>>>>

To prove this result, we will need the following observation.
Note that throughout the proof, all pipes, rows, and column are indexed by indices ranging from $0$ to~$n$.

@@

\begin{proof}
Let~$r$ be the row and $c$ be the column of the crossing~$x$. % (careful that all indices range from~$0$ to~$n$).
Since $i$ has an elbow southwest of $x$, we have $r \le i-1$.
Since $j$ has an elbow northeast of~$x$, there are at most $r-1 \le i-2$ pipes that cross north of $(c,r)$ in column~$c$.
Hence, at least two pipes smaller than~$i$ reach the northern border before column~$c$.
Excluding the pipe~$0$, there is at least one pipe $h > 0$ such that $h < i < j$ and $\omega^{-1}(h) < \omega^{-1}(j) < \omega^{-1}(i)$.
This is a $132$ pattern in the permutation~$\omega$, therefore $\omega$ is not dominant.
\end{proof}

\begin{proof}[Proof of \cref{prob:nuAcyclicProperty}]
For the forward implication, assume that $\omega$ is not dominant, and consider~${i,j,k \in [n]}$ such that~$i < j < k$ while~$\omega^{-1}(i) < \omega^{-1}(k) < \omega^{-1}(j)$, and such that~$k$ is maximal for this property.
Consider any pipe dream~$P \in \pipeDreams(0\omega)$ obtained by capping any pipe dream of~$\pipeDreams(\omega)$ by a top row of contacts.
Because of this top row of contacts, $\omega^{-1}(u) < \omega^{-1}(v)$ implies~$u \contactLess{P} v$.
We now consider the eastmost point~$x$ of pipe~$k$ along row~$i$, and let~$\ell$ denote the other pipe at this point.
We distinguish two cases:
\begin{enumerate}[(i)]
\item Assume first that~$k$ and~$\ell$ cross at~$x$. Note that $k$ is vertical while $\ell$ is horizontal at~$x$. Hence, $\omega^{-1}(k) < \omega^{-1}(\ell)$,  so that~$k \contactLess{P} \ell$. Moreover, $\ell \ne i$ (since~$i$ and~$k$ do not cross), so that~$\ell$ must have an elbow west of~$x$ along row~$i$. As $k$ must have an elbow south of~$x$, we obtain that~$\ell \contactLess{P} k$ by~\cref{lem:rectangle}. Hence, we have both~$k \contactLess{P} \ell$ and~$\ell \contactLess{P} k$, so that~$P$ is cyclic.
\item Assume now that~$k$ and~$\ell$ touch at~$x$. Hence~$k \contactLess{P} \ell$. If~$\ell < k$, then~$k$ and~$\ell$ must cross at~$y$ before~$x$. Flipping~$x$ to~$y$, we obtain a facet~$P'$ with~$\ell \contactLess{P'} k$ (because of the contact~$y$) and~$k \contactLess{P'} \ell$ (because of the last row of elbows which is still in~$P'$), so that~$P'$ would be cyclic. If~$\omega^{-1}(\ell) < \omega^{-1}(k)$, then~$\ell \contactLess{P} k$, so that~$P$ is again cyclic. Hence, we can assume that~$k < \ell$ and~$\omega^{-1}(k) < \omega^{-1}(\ell)$. By maximality of~$k$, we obtain that~$j$ and~$\ell$ do not cross, and thus that~$j$ must have elbow northwest of~$x$. Since $x$ contain an elbow for~$k$, \cref{lem:rectangle} yield~$j \contactLess{P} k$. Since~$\omega^{-1}(k) < \omega^{-1}(j)$, we also have~$k \contactLess{P} j$, so that~$P$ is cyclic.
\end{enumerate}
